---
title: "alluvial-plot"
author: "Erika Braithwaite"
date: '2018-05-02'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(jsonlite, listviewer, tidyr, plotly, ggplot2, magrittr,
               dplyr, alluvial)
URL = 'http://makeup-api.herokuapp.com/api/v1/products.json'
MU = jsonlite::fromJSON(URL) %>% 
        dplyr::select(brand, price, rating, product_type, tag_list)
```

```{r data prep healthy, echo = F}
keep = c('almay', 'annabelle', 'benefit',  'cargo cosmetics', 'clinique', 
         'covergirl', 'dior', 'e.l.f',  "`l'oreal`", 'marcelle', 
         'maybelline', 'nyx', 'physicians formula', 'revlon', 'smashbox', 
         'wet n wild')

complete = MU %>% mutate(tags = as.character(tag_list), 
                         gluten = ifelse(grepl('Gluten', tags), 'gluten_free', 'gluten'), 
                         natural = ifelse(grepl('Natural', tags), 'natural', 'non-natural'), 
                         canadian = ifelse(grepl('Canadian', tags), 'Canadian', 'non-Canadian')) %>% 
        na.omit() %>% 
        filter(brand %in% keep) %>% 
        data.frame()

complete$tag_list = NULL


# Convert brand, category and product_type into factors
cols = c('brand', 'product_type', 'gluten',  'natural', 'canadian')

complete[cols] <- lapply(complete[cols], factor)

# Count the combinations of all the 'healthy' indicators
healthy.dat = complete %>% select(cols) %>% 
        group_by(gluten, natural, canadian) %>% 
        summarise(count = n()) %>% 
        data.frame()

```

## How healthy is your makeup?

Anyone who knows me knows that I love makeup. It's a strange addiction that I can't quite explain: the artistry, the colors, the smells - I adore it all. So when I randomly stumbled across a [makeup API!](https://makeup-api.herokuapp.com/), I could not help but play with the data! 


Restricting my analysis to some of the most common brands found in Canadian drugstores, I ended up with 194 products from 10 different brands. The data contained tags indicating whether a product was 1) Gluten free, 2) Natural, and 3) Canadian. I used R's [alluvial package](https://cran.r-project.org/web/packages/alluvial/vignettes/alluvial.html) which is useful for plotting categorical data.  


```{r alluvial plot, echo = F}
alluvial(healthy.dat[,1:4], freq=healthy.dat$count,
         col = ifelse(healthy.dat$count== "151", "orange", "grey"),
         border = ifelse(healthy.dat$count == "151", "orange", "grey"),
         cex = 0.7, 
         cex.axis = 0.9, 
         axis_labels = c('Gluten \ncontent', 'Natural', 'Canadian', '# of \nproducts')
)

# plot saved interactively
```


Next I wanted to look at makeup ratings by brand and product type. Since most of the brands were considered low- to mid-range in terms of price, I was curious to know whether there would be of variability across brands. I did not have access to higher end brands like Dior or Guerlain, but my guess is that price is often conflated with quality since it is easy to assume that paying more ensures a better product. I have a good deal of thoughts and feelings about that theory. For an in-depth discussion about the 'value' of high end makeup, please contact me directly at erika[at]precision-analytics.ca. 

I plotted ratings against price in order to see how brands compared using ggplotly. Splitting the plot area into quadrants gives me a rough sense of how brands cluster together. 

```{r data prep ratings, echo = F}
rating = MU %>% select(brand, price, rating) %>% 
        filter(brand %in% keep) %>% 
        na.omit() %>% 
        arrange(brand) %>% 
        mutate(price = as.numeric(price), 
               brand = as.factor(brand)) %>% 
        data.frame()
```


```{r rating-price plot, echo = F}
rating.plot = ggplot(data = rating, aes(x = price, y = rating,  colour = brand)) +
        geom_point(alpha = 0.6) + 
        geom_hline(yintercept = 3.5) + 
        geom_vline(xintercept = 20) + 
        theme(legend.position = 'none') + 
        labs(x = 'Price ($)', 
             y = 'Rating (1-5)') + 
        geom_label(aes(label = brand)) + 
        scale_color_brewer(palette="Paired") + 
        theme_bw() + 
        ggtitle('Price and ratings of makeup brands') + 
        theme(legend.title=element_blank())
```





